Follow-up Samples

### 2017-09-13 submission/manuscript_R2.tex
complete modifications to align with back end

\textbf{\emph{The treatment of lesions are not consistent across
diagnosis groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value
\textgreater{} 0.05). It is likely that high inter-individual variation
and the irregular distribution of OTUs across individuals limited the
statistical power of the test. We attempted to overcome these problems
by using Random Forest models to identify collections of OTUs that would
allow us to differentiate between pre and post-treatment samples from
each of the diagnosis groups. To limit the likelihood that the models
would over-fit the data because of the relatively small number of
subjects in each group, we restricted our models to only incorporate 10
OTUs. Despite this restriction, the models performed well (adenoma AUC
range = 0.69 - 0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma
AUC range = 0.82 - 0.98). Interestingly, the 10 OTUs that were used for
each model had little overlap with each other {[}Figure 2{]}. Although
treatment had an impact on the overall community structure, the effect
of treatment was not consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma more
closely resemble those of a normal colon.}} Next, we determined whether
treatment changed the microbiota in a way that the post-treatment
communities resembled that of patients with normal colons. To test this,
we used an expanded cohort of 423 individuals that were diagnosed under
the same protocol as having normal colons or colons with adenoma,
advanced adenoma, or carcinoma {[}Table 2{]}. We then constructed Random
Forest models to classify the study samples, with the 3 diagnosis groups
(adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
models performed well (adenoma AUC range = 0.62 - 0.72, advanced adenoma
AUC range = 0.68 - 0.77, carcinoma AUC range = 0.84 - 0.90; Figure S1).
The OTUs that were incorporated into the adenoma and advanced adenoma
models largely overlapped and those OTUs that were used to classify the
carcinoma samples were largely distinct from those of the other two
models {[}Figure 3A{]}. Among the OTUs that were shared across the three
models were those populations generally considered beneficial to their
host (e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
\emph{Roseburia}) {[}Figures 3B{]}. Although many of these OTUs were
also included in the model differentiating between patients with normal
colons and those with carcinoma, this model also included OTUs
affiliated with populations that have previously been associated with
carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
showing a marked decrease in relative abundance {[}Figure S3{]}.
Finally, we applied these three models to the pre and post-treatment
samples for each diagnosis group and quantified the change in the
positive probability of the model. A decrease in the positive
probability would indicate that the microbiota more closely resembled
that of a patient with a normal colon. There was no significant change
in the positive probability for the adenoma or advanced adenoma groups
(P-value \textgreater{} 0.05) {[}Figure 4{]}. The positive probability
for the pre and post-treatment samples from patients diagnosed with
carcinoma significantly decreased with treatment, suggesting a shift
toward a normal microbiota for most individuals (P-value = 0.02). Only,
6 of the 26 patients (23.08\%) who were diagnosed with a carcinoma had a
higher positive probability after treatment; one of those was
re-diagnosed with carcinoma on the follow up visit. These results
indicate that, although there were changes in the microbiota associated
with treatment, those experienced by patients with carcinoma after
treatment yielded gut bacterial communities of greater similarity to
that of a normal colon.

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). Regardless of
treatment used there was no significant difference in the effect of
these treatments on the number of observed OTUs, Shannon diversity, or
Shannon evenness (P-value \textgreater{} 0.05). Furthermore, there was
not a significant difference in the effect of the treatments on the
amount of change in the community structure (P-value = 0.532). Finally,
the change in the positive probability was not significantly different
between any of the treatment groups (P-value = 0.532). Due to the
relatively small number of samples in each treatment group, it was
difficult to make a definitive statement regarding the specific type of
treatment on the amount of change in the structure of the microbiota.

@@

There was a subset of patients (6 of the 26 with carcinomas) who
demonstrated an elevated probability of carcinoma after treatment. This
may reflect an elevated risk of recurrence. The 23.08\% prevalence of
increased carcinoma probability from our study is within the expected
rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
individuals may have had more severe tumors; however, the tumor severity
of these 6 individuals (3 with Stage II and 3 with Stage III) was
similar to the distribution observed among the other 20 patients. We
also hypothesized that we may have sampled these patients later than the
rest and their communities may have reverted to a carcinoma-associated
state; however, there was not a statistically significant difference in
the length of time between sample collection among those whose
probabilities increased (358 (336 - 458) days) or decreased (334 (256 -
399) days) (Wilcoxon Test; P-value = 0.56) (all days data displayed as
median (IQR)). Finally, it is possible that these patients may not have
responded to treatment as well as the other 20 patients diagnosed with
carcinoma and so the microbiota may not have been impacted the same way.
Again, further studies looking at the role of the microbiota in
recurrence are needed to understand the dynamics following treatment.

@@

All models included only OTU data obtained from 16S rRNA sequencing and
were processed and cleaned using the R package caret (v6.0.76).
Optimization of the mtry hyper-parameter involved making 100 different
80/20 (train/test) splits of the data where the same proportion was
present within both the whole data set and the 80/20 split. For each of
the different splits, 20 repeated 10-fold cross validation was performed
on the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. For all pre versus post-treatment models the
optimized mtry was 2 and for all normal versus diagnosis models the
optimized mtry was 2. The hyper-parameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
was created with the Random Forest model {[}33{]}.

For each of the pre versus post-treatment models assessment of the most
important OTUs was then made by taking the top 10 OTUs by mean decrease
in accuracy (MDA). These were then used to build each respective reduced
OTU pre versus post-treatment model by diagnosis group to help avoid
model over-fitting. These reduced models were then put through the same
process mentioned in the previous paragraph and were what was used for
the final pre versus post-treatment models. For the normal versus
diagnosis models the important OTUs were obtained by counting the number
of times an OTU was present in the top 10\% of MDA for each of the 100
different splits. This was then followed with filtering of this list to
variables that were only present in more than 50\% of these 100 runs.
These corresponding reduced OTU normal versus diagnosis models were then
put through the same process mentioned in the previous paragraph and
were what was used for the final normal versus diagnosis models. For the
pre versus post-treatment models the final optimized mtry was 2 and for
the normal versus diagnosis models the final optimized mtry was 2.

Each model was then applied to our 67-person cohort {[}Table 1{]} based
on diagnosis: adenoma (pre-treatment adenoma (adenoma n = 22 and disease
free n = 0) versus post-treatment adenoma (adenoma n = 0 and disease
free n = 22)), advanced adenoma pre-treatment advanced adenoma (advanced
adenoma n = 19 and disease free n = 0 ) versus post-treatment advanced
adenoma (advanced adenoma n = 0 and disease free n = 19), and carcinoma
(pre-treatment carcinoma (carcinoma n = 26 and disease free n = 0)
versus post-treatment carcinoma (carcinoma n = 1 and disease free n =
25)). The application of the pre versus post-treatment models generated
the probabilities that the sample was a pre-treatment sample. The
application of the normal versus diagnosis models generated the
probabilities that the sample was that specific diagnosis (adenoma,
advanced adenoma, or carcinoma).

@@

\textbf{\emph{Reproducible Methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
have been deposited into the NCBI Sequence Read Archive (SRP062005 and
SRP096978) and the necessary metadata can be found at
\url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the
respective SRA study accession.

@@

\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10
MDA value obtained from 100 different 80/20 splits while the lighter
colored circles represents the value obtained for a specific run.

\textbf{Figure 3: OTUs common to those models used to differentiate
between patients with normal colons and those with adenoma, advanced
adenoma, and carcinoma.} A) Venn diagram showing the OTU overlap between
each model. B) For each common OTU the lowest taxonomic identification
and importance rank for each model run is shown.

@@

\textbf{Figure S2: Summary of important OTUs for the adenoma, advanced
adenoma, and carcinoma models.} A) MDA of the most important variables
in the adenoma model. The dark green point represents the mean and the
lighter green points are the value of each of the 100 different runs. B)
Summary of Important Variables in the advanced adenoma model. MDA of the
most important variables in the SRN model. The dark yellow point
represents the mean and the lighter yellow points are the value of each
of the 100 different runs. C) MDA of the most important variables in the
carcinoma model. The dark red point represents the mean and the lighter
red points are the value of each of the 100 different runs.

>>>>

\textbf{\emph{The treatment of lesions are not consistent across
diagnosis groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value
\textgreater{} 0.05). It is likely that high inter-individual variation
and the irregular distribution of OTUs across individuals limited the
statistical power of the test. We attempted to overcome these problems
by using Random Forest models to identify collections of OTUs that would
allow us to differentiate between pre and post-treatment samples from
each of the diagnosis groups. The models overall performed moderately,
with the carcinoma model being the best (adenoma AUC range = 0.54 -
0.83, advanced adenoma AUC range = 0.34 - 0.65, carcinoma AUC range =
0.82 - 0.98). Interestingly, the top 10 most important OTUs by MDA that
were used for each model had little overlap with each other {[}Figure
2{]}. Although treatment had an impact on the overall community
structure, the effect of treatment was not consistent across patients
and diagnosis groups. Both the adenoma and carcioma treatment models had
AUCs that were significantly higher than a random model permutation
(P-value \textless{} 0.0001).

\textbf{\emph{Post-treatment samples from patients with carcinoma more
closely resemble those of a normal colon.}} Next, we determined whether
treatment changed the microbiota in a way that the post-treatment
communities resembled that of patients with normal colons. To test this,
we used an expanded cohort of 423 individuals that were diagnosed under
the same protocol as having normal colons or colons with adenoma,
advanced adenoma, or carcinoma {[}Table 2{]}. We then constructed Random
Forest models to classify the study samples, with the 3 diagnosis groups
(adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
models performed moderately with CRC being the best (adenoma AUC range =
0.50 - 0.62, advanced adenoma AUC range = 0.53 - 0.67, carcinoma AUC
range = 0.71 - 0.82; Figure S1). The OTUs that were in the top 10\% of
importance for the adenoma and advanced adenoma models largely
overlapped and those OTUs that were used to classify the carcinoma
samples were largely distinct from those of the other two models
{[}Figure 3A{]}. Among the OTUs that were shared across the three models
were those populations generally considered beneficial to their host
(e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
\emph{Roseburia}) {[}Figures 3B{]}. Although many of important OTUs in
the top 10\% were also included in the model differentiating between
patients with normal colons and those with carcinoma, this model also
included OTUs affiliated with populations that have previously been
associated with carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
showing a marked decrease in relative abundance {[}Figure S3{]}.
Finally, we applied these three models to the pre and post-treatment
samples for each diagnosis group and quantified the change in the
positive probability of the model. A decrease in the positive
probability would indicate that the microbiota more closely resembled
that of a patient with a normal colon. There was no significant change
in the positive probability for the adenoma or advanced adenoma groups
(P-value \textgreater{} 0.05) {[}Figure 4{]}. The positive probability
for the pre and post-treatment samples from patients diagnosed with
carcinoma significantly decreased with treatment, suggesting a shift
toward a normal microbiota for most individuals (P-value = 0.001). Only,
7 of the 26 patients (26.92\%) who were diagnosed with a carcinoma had a
higher positive probability after treatment; one of those was
re-diagnosed with carcinoma on the follow up visit. These results
indicate that, although there were changes in the microbiota associated
with treatment, those experienced by patients with carcinoma after
treatment yielded gut bacterial communities of greater similarity to
that of a normal colon.

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). Regardless of
treatment used there was no significant difference in the effect of
these treatments on the number of observed OTUs, Shannon diversity, or
Shannon evenness (P-value \textgreater{} 0.05). Furthermore, there was
not a significant difference in the effect of the treatments on the
amount of change in the community structure (P-value = 0.375). Finally,
the change in the positive probability was not significantly different
between any of the treatment groups (P-value = 0.375). Due to the
relatively small number of samples in each treatment group, it was
difficult to make a definitive statement regarding the specific type of
treatment on the amount of change in the structure of the microbiota.

@@

There was a subset of patients (7 of the 26 with carcinomas) who
demonstrated an elevated probability of carcinoma after treatment. This
may reflect an elevated risk of recurrence. The 26.92\% prevalence of
increased carcinoma probability from our study is within the expected
rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
individuals may have had more severe tumors; however, the tumor severity
of these 7 individuals (1 with Stage I, 3 with Stage II, and 3 with
Stage III) was similar to the distribution observed among the other 19
patients. We also hypothesized that we may have sampled these patients
later than the rest and their communities may have reverted to a
carcinoma-associated state; however, there was not a statistically
significant difference in the length of time between sample collection
among those whose probabilities increased (331 (246 - 358) days) or
decreased (364 (301 - 434) days) (Wilcoxon Test; P-value = 0.39) (all
days data displayed as median (IQR)). Finally, it is possible that these
patients may not have responded to treatment as well as the other 19
patients diagnosed with carcinoma and so the microbiota may not have
been impacted the same way. Again, further studies looking at the role
of the microbiota in recurrence are needed to understand the dynamics
following treatment.

@@

All models included only OTU data obtained from 16S rRNA sequencing and
were processed and cleaned using the R package caret (v6.0.76).
Optimization of the mtry hyper-parameter involved making 100 different
80/20 (train/test) splits of the data where the same proportion was
present within both the whole data set and the 80/20 split. For each of
the different splits, 20 repeated 10-fold cross validation was performed
on the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. For adenoma pre versus post-treatment models the
optimized mtrys were 2 (n = 42), 42 (n = 26), and 914 (n = 32) for
adenoma. For advanced adenoma the mtrys were 2 (n = 26), 52 (n = 14),
and 1353 (n = 60). For carcinoma the mtrys were 2 (n = 36), 41 (n = 54),
and 863 (n = 10). For the normal versus diagnosis models the optimized
mtrys were (n = NA), (n = NA), and (n = NA) for adenoma. For advanced
adenoma (n = NA), (n = NA), and (n = NA). For carcinoma (n = NA), (n =
NA), and (n = NA). The hyper-parameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
was created with the Random Forest model {[}33{]}.

The treatment models were then assessed for their strength to classify
whether individuals had obtained treatment. We also assessed whether the
models performed better than random by creating 100 permutations of a
data set in which the labels were randomly assigned. To assess
similarity of the treatment models we looked for overlap between their
top 10 most important OTUs.

For the diagnosis models that were trained on a completely independent
sample set, each model was then applied to our 67-person cohort {[}Table
1{]} based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = 22
and disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
disease free n = 22)), advanced adenoma pre-treatment advanced adenoma
(advanced adenoma n = 19 and disease free n = 0 ) versus post-treatment
advanced adenoma (advanced adenoma n = 0 and disease free n = 19), and
carcinoma (pre-treatment carcinoma (carcinoma n = 26 and disease free n
= 0) versus post-treatment carcinoma (carcinoma n = 1 and disease free n
= 25)). The application of the pre versus post-treatment models
generated the probabilities that the sample was a pre-treatment sample.
The application of the normal versus diagnosis models generated the
probabilities that the sample was that specific diagnosis (adenoma,
advanced adenoma, or carcinoma). We also looked at similarities between
the models by looking for overlap in the top 10\% of important OTUs to
each respective model.

@@

\textbf{\emph{Reproducible Methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_FollowUps_Microbiome_2017}. Raw
sequences have been deposited into the NCBI Sequence Read Archive
(SRP062005 and SRP096978) and the necessary metadata can be found at
\url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the
respective SRA study accession.

@@

\textbf{Figure 2: The top 10 most important OTUs used to classify
treatment for adenoma, advanced adenoma, and carcinoma.} A) Adenoma
OTUs. B) Advanced Adenoma OTUs. C) Carcinoma OTUs. The darker circle
highlights the median log10 MDA value obtained from 100 different 80/20
splits while the lighter colored circles represents the value obtained
for a specific run.

\textbf{Figure 3: Top 10\% most important OTUs common to those models
used to differentiate between patients with normal colons and those with
adenoma, advanced adenoma, and carcinoma.} A) Venn diagram showing the
OTU overlap between each model. B) For each common OTU the lowest
taxonomic identification and importance rank for each model run is
shown.

@@

\textbf{Figure S2: Summary of top 10\% of important OTUs for the
adenoma, advanced adenoma, and carcinoma models.} A) MDA of the most
important variables in the adenoma model. The dark green point
represents the mean and the lighter green points are the value of each
of the 100 different runs. B) Summary of Important Variables in the
advanced adenoma model. MDA of the most important variables in the SRN
model. The dark yellow point represents the mean and the lighter yellow
points are the value of each of the 100 different runs. C) MDA of the
most important variables in the carcinoma model. The dark red point
represents the mean and the lighter red points are the value of each of
the 100 different runs.
